1. modify GDSArray-class.R with `GDSArray(frompkg=c("SNPRelate", "SeqArray"). )`
2. modify `.write_gdsdata_sampleInCol` for `names(get.attr.gdsn())`. Not `snp.order` or `sample.order` for SeqVarGDSClass data. 
3. SummarizedExperiment(assay=gdsa, colData=colData, rowRanges= rowRange)

4. accessor `gdsfile` for GSE object. S4Vectors)

GDS based SE object, coerce to VRanges. 

```{r}
setGeneric("gdsfile", function(x) standardGeneric("gdsfile"))

setMethod("gdsfile", "GDSArraySeed", function(x) x@file)

setMethod("gdsfile", "GDSArray", function(x) gdsfile(seed(x)))
setMethod("gdsfile", "DelayedArray", function(x) gdsfile(seed(x)))

setMethod("gdsfile", "SummarizedExperiment", function(x) {
    vapply(assays(gse1), gdsfile, character(1))
})
```
import(SNPRelate)
import(VariantAnnotation)
import(gdsfmt)

```{r}
file <- SNPRelate::snpgdsExampleFileName()
gdsa <- GDSArray(file)
f <- snpgdsOpen(file)
colData <- .colData_snpgds(f)
rowRange <- .rowRanges_snpgds(f)
SummarizedExperiment(assay=gdsa, colData=colData, rowRanges= rowRange)
## Error in FUN(X[[i]], ...) : 
##   assay rownames() must be NULL or equal rowData rownames() / rowRanges
##   names()
head(rownames(gdsa))  ## integer
head(names(rowRange))  ## character
```

## send to Martin: 
gds file path
gse1.Rds


data sharing in same location (shared dir on server)
